The Hydrogen Epoch of Reionization Array Dish I: Beam Pattern Measurements and Science Implications

### 2016-02-03 DishBeamMeasurements/paper/DishBeamMeasurementsPaper_rev3.tex
added dish figure, working on power pattern measurement text updates

This is the first in a series of four papers detailing the HERA element. In this work we study 
angular response of the dish and its implications for power spectrum measurements. The three companion 
papers present reflectometry measurements (Patra et al., submitted) and simulations (Ewall-Wice et al., 
submitted) of the dish frequency response, as well as detailed 
foreground simulations for HERA (Thyagarajan et al., submitted). A general description to the design of the 
HERA experiment is given by DeBoer et al. (in prep). In essence, we 
require a large collecting area for
 sensitivity and minimal sidelobes and horizon response without incurring the large cost per collecting 
area of very large dishes. A dish is preferred to a large phased array as it has fewer degrees 
of freedom and reduced potential of antenna-to-antenna variation (Neben et al 2015b, 
submitted). The 331 dishes are positioned in a compact, hexagonal 
array permitting redundant baseline calibration and coherent integration in $
\vec{k}$ space \citep{omniscope,parsons12a}. 
%The smaller field of view of the HERA dish relative to that of the PAPER dipole or the MWA tile does not limit sensitivity in Fourier space as most leverage on high $k$ modes is from line of sight modes anyway \citep{PoberNextGen}.

@@

We review here the design of the HERA dish, and refer to DeBoer et al. (in prep) for details. 
The 14\,m HERA dish design is a departure from the large N--small D approach used by 21\,cm 
observatories like PAPER and the MWA. Both observatories are actively pursuing power 
spectrum analyses using several year data runs, but the sheer data volume makes 
characterization and removal of systematics, as well as repeated or complimentary analyses, 
challenging. A larger element was chosen for HERA primarily to reduce correlation and data processing costs.

The HERA element is a 14\,m diameter parabolic dish with $f/D\approx0.32$ and dual-polarized dipole feed (Fig. \ref{fig:feedphoto}) suspended at prime focus, where $f/D$ is the ratio of focal length to diameter. The dish surface is formed by sheets of wire attached to PVC tubes attached to supports at the edge of the dish and to a hub at the base. The feed is suspended with cables attached to three telephone poles around the dish. The dipole stands 17'' below a 78'' diameter circular mesh back plane surrounded by a 30'' deep cylinder. The cylinder is offset 0.5'' from the back plane. The feed is designed to make the beams for both polarizations more similar to each other, and also to taper the dipole beam towards the edge of the dish to mitigate dish-dish coupling.  The nominal dish focus is at 4.5\,m from the dish surface, though numerical dish models including the feed predict the focus is closer to 5.3\,m.

Feed/dish optimization studies are ongoing and these parameters may change in the final HERA array, but in this work we use the feed with these dimensions and study the beam with the feed suspended at different heights. We study the position of the dish focus, the shape of the main lobe, the magnitude of the dish sidelobes, the degree of beam symmetry, the actual dish/feed efficiency (i.e., collecting area), and the expected level of antenna-to-antenna variation.

@@

One cost of using fewer large antenna elements is a smaller field of view and thus fewer measured Fourier modes perpendicular to the line of sight. 
However, this is a 
small effect for 21\,cm power spectrum analyses as our leverage on $k$ modes comes primarily from modes along the line of sight (in the frequency dimension). Further, HERA's smaller field of view is actually desirable in that it drastically reduces the magnitude of emission at the edge of the wedge compared to a simple dipole element \citep{nithya15}. A second potential drawback is frequency structure introduced by time domain reflections between the suspended feed detailed by Ewall-Wice et al. (submitted) with simulations and Patra et al. (in prep) with zenith reflectometry measurements. These works demonstrate, though, that the slight frequency structure of the dish is small enough so as not to interfere with EOR science.

\subsection{Dish Modeling}
\label{sec:dishmodels}

\begin{figure*}
\centering
\includegraphics[width=7in]{dave197_rich195_airy_beams.pdf}
\caption{Simulated dish power patterns at 137\,MHz (Sec. \ref{sec:dishmodels}) using HFSS (left) and CST (middle) are shown beside an ideal Airy pattern for a 14\,m diameter dish for comparison.}
\label{fig:modelbeams}
\end{figure*}

We numerically model the HERA dish in both ANSYS HFSS\footnote{http://www.ansys.com/Products/Electronics/ANSYS-HFSS} and in CST Microwave Studio\footnote{https://www.cst.com/Products/CSTMWS}, making slightly different assumptions, in order to study the dependence of the beam pattern on modeling imperfections. In particular, we suspect that the near horizon beam response, which sets the level of horizon brightening, is quite sensitive to modeling assumptions. Modeling is also useful to probe the best focus of the dish--feed system. The nominal focus is at 4.5\,m given that the dish has $f/D=0.32$, though the feed back plane and skirt are expected to move the focus slightly.

The HFSS model (Fig. \ref{fig:modelbeams}, left panel) uses a 14\,m paraboloid with $f/D=0.32$ as the dish, with a 1\,m hole at the vertex. In reality the 1\,m diameter at the vertex is filled with concrete and supports the PVC pipes which suspend the mesh panels, but we effectively assume here it is perfectly absorbant.  The feed is modeled as a PAPER dipole suspended 17'' below a 78'' diameter circular backplane inside a 30'' deep skirt hanging 0.5'' below the backplane. All these surfaces are modeled as solid aluminum, as are the dipole ``sleeves'', circular disks just above and below the dipole plane designed to broaden its frequency response. The dipoles themselves is modeled as a copper pipe of diameter XX. In the simulations, we excite one of the dipoles using a modal port and measure the total gain response in each direction to 137.5\,MHz radiation.

In the CST model (Fig. \ref{fig:modelbeams}, middle panel) we again model the dish as a 14\,m diameter solid aluminum paraboloid with $f/D=0.32$ but now neglect the 1\,m hole at the vertex, effectively assuming the concrete and earth ground behind it are perfectly reflective. The actual material properties are somewhere in between, though difficult to predict theoretically. The feed model has the same dimensions as in the HFSS model. The simulations are done by exciting one of the dipoles with band-limited noise (100--200\,MHz). The beam pattern is then obtained at a given frequency using a farfield monitor after the excitation pulse energy within the structure decays to below -80\,dB. As expected, both models have slightly stronger sidelobes and wider main lobes than the ideal Airy pattern (Fig. \ref{fig:modelbeams}, right panel).

@@

\begin{figure*}[t]
\centering
\includegraphics[width=6.5in]{measured_beams_and_models_maps.pdf}
\caption{Measured dish power patterns at various feed heights for the EW (left panel) and NS (right panel) linear instrumental polarizations. The sidelobes shrink and the main lobe narrows as the feed is raised, confirming that the focus of the dish/feed system is higher than the 4.5\,m ideal focus of the parabolic dish.}
\label{fig:measuredbeammaps}
\end{figure*}

\begin{figure*}[t]
\centering
\includegraphics[width=6.5in]{measured_beams_and_models_slices.pdf}
\caption{Slices through the E and H planes through the measured dish power patterns (points) and numerical models (curves). The measured beams agree with both models in the main lobe out to zenith angles of 15--20$^\circ$, but begin to deviate in the sidelobes at the 0.1--0.5\% level of the beam.}
\label{fig:measuredbeamslices}
\end{figure*}

We make three dish power pattern measurements with the feed at different heights: (1) 4.5\,m, the nominal dish focus; (2) 5.0\,m, an intermediate focus; and (3) 5.3\,m, the numerically determined focus of the dish/feed system, where all heights are measured from the dish surface to the feed back plane. In each configuration we collect data for 2--4 days, obtaining roughly 200 satellite passes. We exclude times when the received power is within 20\,dB of the background level determined between passes, and then grid measured beam values into 1.8$^\circ$ HEALPix cells on the sky, rejecting outliers in the top or bottom 5\% in each cell as a final guard against rare satellite identification problems or ADC saturation issues.

@@

 \begin{table}[h]
 \caption{ \label{table:collectingareatable}Collecting area (m$^2$) of measured 137\,MHz beams and corresponding power spectrum SNR for HERA-331.}
\begin{tabular}{| l | l | l |}
\hline
Beam & $A_\text{eff}$ (m$^2$) & SNR (pess, mod, opt)\\
\hline
  Airy pattern & 155 & 21.4, 25.5, 90.8  \\
    Measured, feed at 5.3\,m & 97.9 & 18.7, 19.3, 74.3 \\
    Measured, feed at 5\,m & 82.6 & 15.9, 16.4, 67.9 \\
    Measured, feed at 4.5\,m & 73.6 & 15.1, 15.4, 63.9 \\ 
  \hline
\end{tabular}
\end{table}

% To input these collecting areas into 21cmSense, we convert these measured dish collecting areas into effective dish diameters, which we input as the \texttt{dish\_size\_in\_lambda} parameter. 
We run 21cmSense\footnote{https://github.com/jpober/21cmSense} to compute the overall SNR of a power spectrum detection with one season (6 hours per night for 180 nights) of HERA-331 data.We predict the SNRs with optimistic, moderate, and pessimistic foreground assumptions. In the optimistic case, $k$ modes inside the same $uv$ pixel are added coherently, and all modes whose frequency dependence is larger than that of source at the edge of the main lobe are used. In the moderate case, $k$ modes inside the same $uv$ pixel are added coherently and only modes whose frequency dependence falls outside of the horizon plus a buffer are used. In the pessimistic case, all baselines are added incoherently and only modes outside the horizon plus a buffer are used.

@@

@@

We predict the magnitude of this effect for the HERA element as well as its dependence on beam modeling uncertainties. As expected, we find that the level of horizon brightening is largest when the galaxy is just above the horizon, and lowest when it is well below. When the pitchfork effect is large, its uncertainty is also large as seen in the deviation of the delay spectra with HFSS and CST beam models in these simulations. When the effect is small, the two beam models produce much more similar results, highlighting the delay spectrum as an exquisite probe of the challenging to measure beam response at very low elevations.

% frequency dependence of the beam pattern and overall gain (due to reflections) likely dominates leakage deep within the window, but the other papers in this series demonstrate that this leakage level is within our spec

As discussed by the other papers in this series, the frequency dependence of the beam power pattern and its overall gain widen the delay kernel of a source, leaking power into the EOR window. These papers demonstrate that in the best case, a combination of bandpass calibration and delay space CLEANing can mitigate foregrounds to a manageable level, while in the worst case, HERA can still achieve a XX--XX $\sigma$ detection of the EOR parameters from the power spectrum by simply excluding a very conservative buffer outside the wedge. All these considerations highlight the need for precision characterization of antenna elements and their implications for 21\,cm science analyses on the part of next generation observatories.

>>>>

This is the first in a series of four papers detailing the HERA element. In this work we study 
angular response of the dish and its implications for power spectrum measurements. The three companion 
papers present reflectometry measurements (Patra et al., submitted) and simulations (Ewall-Wice et al., 
submitted) of the dish frequency response, as well as detailed 
foreground simulations for HERA (Thyagarajan et al., submitted). A general description to the design of the 
HERA experiment is given by DeBoer et al. (in prep). In essence, we 
require a large collecting area for
 sensitivity and minimal sidelobes and horizon response without incurring the large cost per collecting 
area of very large dishes. A dish is preferred to a large phased array as it has a less complex beampattern and reduced potential for antenna-to-antenna variation (Neben et al, submitted). The 331 dishes are positioned in a compact, hexagonal 
array permitting redundant baseline calibration and coherent integration in $
\vec{k}$ space \citep{omniscope,parsons12a}. 
%The smaller field of view of the HERA dish relative to that of the PAPER dipole or the MWA tile does not limit sensitivity in Fourier space as most leverage on high $k$ modes is from line of sight modes anyway \citep{PoberNextGen}.

@@

The HERA element (Fig. \ref{fig:feeddiagram}) is a 14\,m diameter faceted parabolic dish ($f\#=0.32$) with a dual-polarized dipole feed suspended at prime focus. The dish surface is formed by wire mesh sheets (i.e., facets) mounted on PVC tubes which run from the lip of the dish to the hub at the vertex. The feed consists of a dual linear polarization PAPER sleeved dipole mounted 17'' below a 78'' diameter wire mesh back plane surrounded by a 30'' deep cylinder, suspended from cables attached to each the three telephone poles around the dish. The dipole ``sleeves'' are circular disks just above and below the dipole designed to broaden its frequency response. The cylinder is offset 0.5'' from the back plane, and is designed to make the beams for both linear polarizations more similar, and also to taper the dipole beam towards the edge of the dish to mitigate dish-dish coupling. Fig. \ref{fig:feedphoto} shows the feed as deployed on the ground for early testing.  The nominal dish focus is $f=(f\#)D=4.48$\,m, though our numerical electromagnetic models suggest the actual focus is slightly higher due to the feed geometry. In this work we study the dish beam pattern at feed heights of 4.5\,m, 5.0\,m, and 5.3\,m, measured from dish surface to feed plane. These height measurements are uncertain at the $\pm5\%$ level in this study. For more details on the dish design and construction see DeBoer et al. (in prep). Also note that feed/dish optimization studies are ongoing and the values of these parameters may change in the full HERA array.

\begin{figure}[h]
\includegraphics[width=3.4in]{dish_and_feed_diagram.pdf}
\caption{some stuff here}
\label{fig:feeddiagram}
\end{figure}

@@

As the HERA element is larger than the MWA or PAPER antenna elements, one might worry about the  smaller field of view and thus smaller number of measured Fourier modes perpendicular to the line of sight. However, this is a 
small effect for 21\,cm power spectrum analyses as our leverage on $k$ modes comes primarily from modes along the line of sight (in the frequency dimension). Further, HERA's smaller field of view is actually desirable in that it drastically reduces the magnitude of emission at the edge of the wedge compared to a simple dipole element \citep{nithya15}. A second potential drawback is frequency structure introduced by time domain reflections between the suspended feed detailed by Ewall-Wice et al. (submitted) with simulations and Patra et al. (in prep) with zenith reflectometry measurements. These works demonstrate, though, that the slight frequency structure of the dish is sufficiently small to not interfere with EOR science.
\\
\subsection{Dish Modeling}
\label{sec:dishmodels}

\begin{figure*}
\centering
\includegraphics[width=7in]{dave197_rich195_airy_beams.pdf}
\caption{Simulated dish power patterns at 137\,MHz (Sec. \ref{sec:dishmodels}) with $h_\text{feed}=5$\,m using HFSS (left) and CST (middle) are shown beside an ideal Airy pattern for a 14\,m diameter dish for comparison.}
\label{fig:modelbeams}
\end{figure*}

We numerically model the HERA dish in both ANSYS HFSS\footnote{http://www.ansys.com/Products/Electronics/ANSYS-HFSS} and in CST Microwave Studio\footnote{https://www.cst.com/Products/CSTMWS}, making slightly different assumptions, in order to study the dependence of the beam pattern on modeling imperfections. In particular, the near horizon beam response, which sets the level of horizon brightening in the delay spectrum, is quite sensitive to modeling assumptions. Modeling also probes the best focus of the dish--feed system.

The HFSS beam model (Fig. \ref{fig:modelbeams}, left panel) assumes a solid 14\,m parabolic dish ($f\#=0.32$) with a 1\,m hole at the vertex. In reality a 1\,m diameter circle at the vertex is filled with concrete and supports the PVC pipes which suspend the mesh panels, but we effectively assume here it is perfectly absorbant.  The feed is modeled as sleeved dipole with a solid metal back plane and cylinder with the dimensions given above. Copper is used for the dipole itself, and aluminum for all reflecting surfaces. In the simulations, we excite one of the dipoles using a modal port and measure the total gain response in each direction to 137.5\,MHz radiation.

The CST beam model (Fig. \ref{fig:modelbeams}, middle panel) again assumes a 14\,m parabolic diameter solid aluminum paraboloid with $f/D=0.32$ but neglects the 1\,m hole at the vertex, effectively assuming the concrete and earth ground behind it are perfectly reflective. The actual material properties are somewhere between fully reflective and fully absorbant, and thus, between these two models. The feed model has the same dimensions and properties as in the HFSS model. These simulations are done by exciting one of the dipoles with band-limited noise (100--200\,MHz), then measuring the farfield radiation in each direction after the excitation pulse energy within the structure decays to below -80\,dB. As expected, both models have slightly stronger sidelobes and wider main lobes than the ideal Airy pattern (Fig. \ref{fig:modelbeams}, right panel). In both models, the best focus is found to be close to 5.3\,m with this feed geometry, slightly larger than the nominal focus of 4.5\,m, as expected.

@@

\begin{figure*}[t]
\centering
\includegraphics[width=6.5in]{measured_beams_and_models_maps.pdf}
\caption{Measured dish power patterns with three feed heights for the EW (left panel) and NS (right panel) linear instrumental polarizations. The sidelobes shrink and the main lobe narrows as the feed is raised, confirming that the best focus of the dish/feed system is close to $h_\text{feed}=5.3$\,m.}
\label{fig:measuredbeammaps}
\end{figure*}

\begin{figure*}[t]
\centering
\includegraphics[width=6.5in]{measured_beams_and_models_slices.pdf}
\caption{Slices through the E and H planes through the measured dish power patterns (points) and numerical models (curves). The measured beams agree with both models in the main lobe out to zenith angles of 15--20$^\circ$, but begin to deviate in the sidelobes where the beam response is 25-30\,dB down from zenith.}
\label{fig:measuredbeamslices}
\end{figure*}

feed misalignment vs surface imperfections (different across different feed heights)
risk of cross-coupling
tie-downs

We make dish power pattern measurements with the feed at 4.5\,m, 5.0\,m, and 5.3\,m height above the dish surface (see Fig. \ref{fig:feeddiagram}).

In each configuration we collect data for 2--4 days, obtaining roughly 200 satellite passes. We exclude times when the received power is within 20\,dB of the background level determined between passes, and then grid measured beam values into 1.8$^\circ$ HEALPix cells on the sky, rejecting outliers in the top or bottom 5\% in each cell as a final guard against rare satellite identification problems or ADC saturation issues.

@@

 \begin{table}[h]
 \caption{ \label{table:collectingareatable}Collecting area (m$^2$) of measured 137\,MHz beams and corresponding power spectrum SNR for HERA-331.}
\begin{tabular}{| l | l | l |}
\hline
Beam & $A_\text{eff}$ (m$^2$) & SNR (pess, mod, opt)\\
\hline
  Airy pattern & 155 & 21.4, 25.5, 90.8  \\
    Measured, feed at 5.3\,m & 93.0 & 18.7, 19.3, 74.3 \\
    Measured, feed at 5\,m & 77.1 & 15.9, 16.4, 67.9 \\
    Measured, feed at 4.5\,m & 68.5 & 15.1, 15.4, 63.9 \\ 
  \hline
\end{tabular}
\end{table}

% To input these collecting areas into 21cmSense, we convert these measured dish collecting areas into effective dish diameters, which we input as the \texttt{dish\_size\_in\_lambda} parameter. 
We run 21cmSense\footnote{https://github.com/jpober/21cmSense} to compute the overall SNR of a power spectrum detection with one season (6 hours per night for 180 nights) of HERA-331 data. We predict the SNRs with optimistic, moderate, and pessimistic foreground assumptions assuming a fiducial Epoch of Reionization model generated with 21cmFast \citep{21cmfast}. This model assumes $\zeta=31.5$ for the ionizing efficiency, $T_\mathrm{vir}=1.5\times10^4$\,K for the minimum virial temperature of halos producing ionizing photons, and $R_\text{mfp}=30$\,Mpc for the mean free path of ionizing photons, and reaches 50\% ionization at $z \sim 9.5$ and complete ionization at $z \sim 7$, and is consistent with most current observations \citep[e.g.][]{PoberNextGen}. In the optimistic case, $k$ modes inside the same $uv$ pixel are added coherently, and all modes whose frequency dependence is larger than that of source at the edge of the main lobe are used. In the moderate case, $k$ modes inside the same $uv$ pixel are added coherently and only modes whose frequency dependence falls outside of the horizon plus a buffer of $\Delta k_\parallel=0.1\,\mathrm{h}\,\mathrm{Mpc}^{-1}$ are used. In the pessimistic case, all baselines are added incoherently and only modes outside the horizon plus the same buffer are used.

@@

As discussed by the other papers in this series, the frequency dependence of both the beam's angular response and its overall gain widen the delay kernel of a source, leaking power into the EOR window out to $k_\parallel=$0.15-0.23h\,Mpc$^{-1}$. This is well inside the wedge plus buffer assumed in the moderate and pessimistic cases considered above. The optimistic case requires calibrating out the instrument frequency response enough to begin to access cosmological modes within the wedge. This is much more challenging, but the sensitivity gains are enormous as those signal modes are thought to be the strongest.

@@

We predict the magnitude of this effect for the HERA element as well as its dependence on beam modeling uncertainties. As expected, we find that the level of horizon brightening is largest when the galaxy is just above the horizon, and lowest when it is well below. When the pitchfork effect is large, its uncertainty is also large as seen in the deviation of the delay spectra with HFSS and CST beam models in these simulations. When the effect is small, the two beam models produce much more similar results, highlighting the delay spectrum as an exquisite probe of the difficult-to-measure beam response at very low elevations.
